\vspace{1.0cm} \LARGE \bf Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex

### 2014-02-19 fluB.tex
A few edits/comments

Influenza B viruses are increasingly being recognized as major contributors to morbidity attributed to seasonal influenza. 
Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages referred to as B/Victoria and B/Yamagata lineages. 
Frequent reassortment between the segments of these two lineages has been noted in the past, but the effects of these reassortments have not been investigated in much detail.
We investigate inter-lineage reassortments by comparing phylogenetic trees across genomic segments.
Our analyses indicate that of the 8 segments of influenza B viruses only PB1, PB2 and HA segments maintained separate Victoria and Yamagata lineages and that currently circulating strains possess PB1, PB2 and HA segments derived entirely from one or the other lineage; other segments have repeatedly reassorted between lineages thereby reducing genetic diversity.
We argue that this difference between segments is due to selection against reassortant viruses with mixed lineage PB1, PB2 and HA segments.
Given sufficient time and continued recruitment to the reassortment-isolated PB1-PB2-HA gene complex, we expect influenza B viruses to eventually undergo sympatric speciation.

@@

\section*{Introduction}
Seasonal influenza causes between 250,000 and 500,000 deaths annually and is comprised of three virus types (A, B and C) co-circulating in humans, of which influenza A is considered to cause the majority of seasonal morbidity and mortality \cite{flufactsheet}.
However, influenza B viruses are increasingly being recognized as important human pathogens \cite{paul-glezen2013}.
Following the 2009 A/H1N1 pandemic, influenza B has increased in prevalence and in the 2012/2013 European season as many as 53\% of influenza sentinel surveillance samples tested positive for influenza B \cite{ECDC1213}.

Like other members of \textit{Orthomyxoviridae}, influenza B viruses have segmented genomes, which allow viruses co-infecting the same cell to exchange segments, a process known as reassortment. 
Influenza A viruses are widely considered to be a major threat to human health worldwide due to their ability to cause pandemics in humans via reassortment of circulating human strains with non-human influenza A strains. 
Influenza B viruses are thought to primarily infect humans (and occasionaly seals \cite{osterhaus2000,bodewes2013} through a reverse zoonosis), and are widely considered to be unable to exhibit pandemics through the acquisition of antigenic novelty from an animal reservoir. 
Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \cite{bedford2014}.

@@

We show that despite extensive reassortment, three of the eight segments -- two segments coding for components of the influenza B virus polymerase, PB1 and PB1, and the surface glycoprotein HA -- still survive as distinct Victoria and Yamagata lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage have rarely been isolated and only circulate as transient lineages once isolated.
In other segments (PA, NP, NA, MP and NS) a single lineage has introgressed into the opposing background and been fixed in the influenza B population: Yam for PA, NP, NA and MP and Vic for NS.
This has occured through repeated reassortments and subsequent fixation of reassortant genome constellations within the influenza B population.

>>>>

Influenza B viruses are increasingly being recognized as major contributors to morbidity attributed to seasonal influenza. 
Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages referred to as B/Victoria and B/Yamagata lineages. 
Frequent exchange of genomic segments of these two lineages has been noted in the past, but the effects %% AR - effects or 
of these reassortments have not been investigated in detail.
We investigate inter-lineage reassortments by comparing phylogenetic trees across genomic segments.
Our analyses indicate that of the 8 segments of influenza B viruses only PB1, PB2 and HA segments maintained separate Victoria and Yamagata lineages and that currently circulating strains possess PB1, PB2 and HA segments derived entirely from one or the other lineage; other segments have repeatedly reassorted between lineages thereby reducing genetic diversity.
We argue that this difference between segments is due to selection against reassortant viruses with mixed lineage PB1, PB2 and HA segments.
Given sufficient time and continued recruitment to the reassortment-isolated PB1-PB2-HA gene complex, we expect influenza B viruses to eventually undergo sympatric speciation.

@@

\section*{Introduction}
Seasonal influenza causes between 250,000 and 500,000 deaths annually and comprises lineages from three virus types (A, B and C) co-circulating in humans, of which influenza A is considered to cause the majority of seasonal morbidity and mortality \cite{flufactsheet}.
However, influenza B viruses are increasingly being recognized as important human pathogens \cite{paul-glezen2013}.
Following the 2009 A/H1N1 pandemic, influenza B has increased in prevalence and in the 2012/2013 European season as many as 53\% of influenza sentinel surveillance samples tested positive for influenza B \cite{ECDC1213}.

Like other members of \textit{Orthomyxoviridae}, influenza B viruses have segmented genomes, which allow viruses co-infecting the same cell to exchange segments, a process known as reassortment. 
Influenza A viruses are widely considered to be a major threat to human health worldwide due to their ability to cause pandemics in humans via reassortment of circulating human strains with non-human influenza A strains. 
Although influenza B viruses have been observed to infect seals \cite{osterhaus2000,bodewes2013} through a reverse zoonosis, they are thought to primarily infect humans and are thus are unlikely to exhibit pandemics due to the absence of an animal reservoir from which to acquire antigenic novelty. 
Both influenza A and B evolve antigenically through time in a process known as antigenic drift, in which mutations to the haemagglutinin (HA) protein allow viruses to escape existing human immunity and persist in the human population, leading to recurrent seasonal epidemics \cite{bedford2014}. %%AR - some earlier references here

@@

We show that despite extensive reassortment, three of the eight segments -- two segments coding for components of the influenza B virus polymerase, PB1 and PB1, and the surface glycoprotein HA -- still survive as distinct Victoria and Yamagata lineages, which appear to be co-adapted to the point where virions which do not contain PB1, PB2 or HA segments derived entirely from either the Vic or the Yam lineage have rarely been isolated and only circulate as transient lineages once isolated.
In other segments (PA, NP, NA, MP and NS) a single lineage has introgressed into the opposing background and been fixed in the influenza B population: Yam for PA, NP, NA and MP and Vic for NS.
This has occurred through repeated reassortments and subsequent fixation of reassortant genome constellations within the influenza B population.
